---
title: "Exploratory Data Analysis"
author: "Srijit Mukherjee"
date: "23 June 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, size = "tiny")
```

## Types of Data Variables in R


- Quantitative

    - Continuous
    - Discrete
      
- Qualitative

    - Nominal
    - Ordinal

## Datasets in R

If you type the following command, you will get to know all the data sets in R.

```{r, echo = TRUE, comment = ''}
data()
library(DAAG)
```

## Understanding Data in

Let's see the top of the data matrix.

```{r, echo = TRUE, comment = ''}
data = ais
head(data, n = 3)
```

## Handling Data in R

Let's see the bottom of the data matrix.

```{r, echo = TRUE, comment = ''}
tail(data, n = 3)
```

## Handling Data in R

Let's see a concise summary of the data matrix.

```{r, echo = TRUE, comment = ''}
str(data)
```

## Handling Data in R

```{r, echo = TRUE, comment = ''}
class(data)
class(data$mpg)
dim(data)
names(data)
```

## Handling Data in R

Exploratory Data Analysis

```{r, echo = TRUE, comment = ''}
summary(data)
```

## Univariate Quantitative Data Analysis

Exploratory Data Analysis

```{r, echo = TRUE, comment = ''}
mean(data$hg)
var(data$hg)
sd(data$hg)
min(data$hg)
```


## Univariate Quantitative Data Analysis

Exploratory Data Analysis

```{r, echo = TRUE, comment = ''}
max(data$hg)
median(data$hg)
quantile(data$hg)
range(data$hg)
```

## Univariate Quantitative Data Analysis

Summary of the Data

```{r, echo = TRUE, comment = ''}
summary(data$hg)
```


## Univariate Quantitative Data Analysis

Histogram

```{r, echo = TRUE, comment = ''}
hist(data$hg, xlab = "Hemaglobin concentration", probability = TRUE, , main = "Histogram of Hemaglobin concentration")
```


## Univariate Quantitative Data Analysis

Kernel Density

```{r, echo = TRUE, comment = ''}
d <- density(data$hg)
plot(d, main = "Kernel density of Hemaglobin concentration")
polygon(d, col = "red", border = "blue")
```

## Univariate Quantitative Data Analysis

Box Plot

```{r, echo = TRUE, comment = ''}
boxplot(data$hg,
        main = toupper("Boxplot of Hemaglobin concentration"),
        ylab = "Hemaglobin concentration",
        col = "red")
```

## Univariate Qualitative Data Analysis

Frequency Distribution Table  

```{r, echo = TRUE, comment = ''}
table(data$sport)
```

## Univariate Qualitative Data Analysis

Vertical Bar Plot

```{r, echo = TRUE, comment = ''}
barplot(table(data$sport), main="Count of participants in different sports for study", xlab="Sports",ylab="Count", border="red", col="blue", density=10)
```


## Univariate Qualitative Data Analysis

Horizontal Bar Plot

```{r, echo = TRUE, comment = ''}
barplot(table(data$sport), main="Count of participants in different sports for study", xlab="Sports",ylab="Count", col = "darkred", horiz = TRUE)
```

## Univariate Qualitative Data Analysis

How to do probability bar plot?

Pie Chart

```{r, echo = TRUE, comment = ''}
pie(table(data$sport), labels = levels(data$sport))
```


## Multivariate Exploratory Data Analysis

## Categorical vs Categorical

Contigency Table

```{r, echo = TRUE, comment = ''}
sex_vs_sport = data[,12:13]
table(sex_vs_sport)
#xtabs(~ sex + sport, sex_vs_sport)
```


## Categorical vs Categorical

Bar Plots Vertical Comparison

```{r, echo = TRUE, comment = ''}
barplot(table(sex_vs_sport),
        main = "Sports Participation Distribution by Sex",
        xlab = "Sport",
        col = c("red","green")
        
)
legend("topleft",
       c("Female","Male"),
       fill = c("red","green")
)
```


## Categorical vs Categorical


Bar Plot Beside Comparison

```{r, echo = TRUE, comment = ''}
barplot(table(sex_vs_sport),
        main = "Sports Participation Distribution by Sex",
        xlab = "Sport",
        col = c("red","green"),
        beside =  TRUE
)
legend("topleft",
       c("Female","Male"),
       fill = c("red","green")
)
```


## Categorical vs Categorical


Pie Chart

```{r, echo = TRUE, comment = ''}
pie(table(data$sport), labels = levels(data$sport))
```


## Continuous vs Categorical

Comparison of Box Plot

```{r, echo = TRUE, comment = ''}
boxplot(hg~sport,
        data=data,
        main="Different boxplots for each sport",
        xlab="Sport",
        ylab="Hemaglobin concentration",
        col="orange",
        border="brown"
)
```



## Continuous vs Categorical

Comparison of Histogram

```{r, echo = TRUE, comment = ''}
library(ggplot2)

hg_vs_sport = data[,c(4,13)]
ggplot(hg_vs_sport, aes(hg, fill = sport)) + geom_density(alpha = 0.2)

```


## Continuous vs Categorical

Comparison of Histogram

```{r, echo = TRUE, comment = ''}
library(ggplot2)

hg_vs_sex = data[,c(4,12)]
ggplot(hg_vs_sex, aes(hg, fill = sex)) + geom_density(alpha = 0.2)

```


## Continuous vs Categorical

 Comparing Summary Data
 
```{r, echo = TRUE, comment = ''}
by(hg_vs_sex, hg_vs_sex$sex, summary)
```

## Continuous vs Continuous

Plot
 
```{r, echo = TRUE, comment = ''}
plot(data$wt ~ data$ht , data,
            xlab="Weight", ylab="Height",
            main="Scatter Plot")
```

## Continuous vs Continuous

Scatter Plot
 
```{r, echo = TRUE, comment = ''}
library(car)
scatterplot(data$wt ~ data$ht , data,
            xlab="Weight", ylab="Height",
            main="Enhanced Scatter Plot")
```

